---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mc3logit: Matched Case-Control Conditional Logit

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/sima-njf/mc3logit_project/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sima-njf/mc3logit_project/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/sima-njf/mc3logit_project/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/sima-njf/mc3logit_project/actions/workflows/test-coverage.yaml)
[![pkgdown](https://github.com/sima-njf/mc3logit_project/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/sima-njf/mc3logit_project/actions/workflows/pkgdown.yaml)
[![codecov](https://codecov.io/gh/sima-njf/mc3logit_project/graph/badge.svg)](https://app.codecov.io/gh/sima-njf/mc3logit_project)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://github.com/sima-njf/mc3logit_project/blob/main/LICENSE.md)
<!-- badges: end -->

The `mc3logit` package implements permutation-based inference for conditional
logistic regression models fit to **matched case-control data** -- the kind
of data you get when several individuals share the same event and only some
of them "act" (e.g. several police officers responding to the same incident,
only some of whom draw their firearm). It builds on `survival::clogit()` and
adds permutation-based p-values and confidence intervals, following the
approach used by Ridgeway (2016) for exactly this kind of matched-officer
data.

The package also ships:

- Two data simulators, `sim_events()` and `sim_events2()`, for generating
  synthetic matched case-control event data (useful for learning the
  package, and for power / type-I-error studies).
- `exposure_dyn()`, for computing running direct/indirect "exposure"
  covariates from a longitudinal event log without double-counting or
  looking into the future.
- `print()`/`plot()`/`confint()` methods for exploring and reporting a
  fitted `clogit_perm` model.

> This package originated in George G. Vega Yon's
> [use_of_force](https://github.com/gvegayon/use_of_force) project (see
> `LICENSE`); this repository packages it standalone with expanded
> vignettes and documentation.

## Installation

<!-- You can install the released version of mc3logit from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("mc3logit") -->
<!-- ``` -->

You can install `mc3logit` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("sima-njf/mc3logit_project", build_vignettes = TRUE)
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r sim-fit, cache=TRUE}
library(mc3logit)

# Simulating data
x <- sim_events(200, 300, seed = 122)

# Fitting
ans <- clogit_perm(
  nperm = 1000,
  pointed000001 ~ female + years + exposed + strata(incidentid),
  data = x
  )
```

```{r print-plot}
print(ans)
plot(ans)
```

## Learning more

Full documentation, including all vignettes, is at
<https://sima-njf.github.io/mc3logit_project/>. The package ships five
vignettes:

- `vignette("getting-started")` -- simulate data, fit `clogit_perm()`, and
  read off `print()`/`plot()`/`confint()` output.
- `vignette("permutation-inference")` -- how within-stratum permutations are
  drawn, and how p-values and both flavors of confidence interval are
  computed.
- `vignette("validation")` -- measured type-I error and power of the
  permutation test, plus a documented limitation (see below).
- `vignette("simulating-data")` -- the `sim_events()`/`sim_events2()`
  data-generating process, tuning effect sizes, and using `nsims`.
- `vignette("dynamic-exposure")` -- computing direct and indirect exposure
  covariates from an event log with `exposure_dyn()`.

Browse them locally with `browseVignettes("mc3logit")` once installed, or
read the source under [`vignettes/`](vignettes/).

## A note on validity

Permuting the raw outcome builds a reference distribution under the
*complete* null -- "no covariate matters" -- rather than the *partial* null
usually of interest, "this covariate doesn't matter, the others may." When
another covariate in the model carries substantial signal, the reference
distribution is too narrow and the resulting p-values are
**anti-conservative**.

`vignette("validation")` measures this directly. Treat `clogit_perm()`
p-values as reliable when the remaining covariates are weak, and cross-check
against the Wald p-values in `$fit` otherwise. Residual-permutation schemes
(Freedman & Lane, 1983) would address it and are not yet implemented.

## References

Ridgeway, G. (2016). Officer risk factors associated with police shootings:
a matched case-control study. *Statistics and Public Policy*, 3(1), 1-6.

Knijnenburg, T. A., Wessels, L. F. A., Reinders, M. J. T., & Shmulevich, I.
(2009). Fewer permutations, more accurate P-values. *Bioinformatics*,
25(12), 161-168.
