---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mc3logit: Matched Case-Control Conditional Logit

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/sima-njf/mc3logit_project/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sima-njf/mc3logit_project/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/sima-njf/mc3logit_project/branch/master/graph/badge.svg)](https://app.codecov.io/gh/sima-njf/mc3logit_project)
<!-- badges: end -->

The `mc3logit` package implements permutation-based inference for conditional logistic regression models. Permutation-based methods for matched case-control logit were notably used by Ridgeway (2016).

## Installation

<!-- You can install the released version of mc3logit from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("mc3logit") -->
<!-- ``` -->

You can install `mc3logit` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("sima-njf/mc3logit_project", subdir = "rpackage")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r sim-fit, cache=TRUE}
library(mc3logit)

# Simulating data
x <- sim_events(200, 300, seed = 122)

# Fitting
ans <- clogit_perm(
  nperm = 1000,
  pointed000001 ~ female + years + exposed + strata(incidentid),
  data = x
  )
```

```{r print-plot}
print(ans)
plot(ans)
```

