---
title: "Computing Dynamic Exposure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing Dynamic Exposure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mc3logit)
```

A recurring covariate in matched case-control studies of contagious
behavior (peer effects, "exposure to violence," etc.) is: *has this
individual been near this behavior before, and how much?* `exposure_dyn()`
computes that kind of running exposure count from a longitudinal event
log, correctly handling two things that are easy to get wrong by hand:

- **Temporal ordering** -- exposure must only look *backwards* in time.
- **No double counting** -- an individual's own action in the current event
  shouldn't count as exposure to themselves.

## The two kinds of exposure

Given a data frame with one row per (individual, event) pair and a binary
`action` indicator, `exposure_dyn()` distinguishes:

- **Direct exposure** -- did *this* individual act in a *previous* event?
- **Indirect exposure** -- did *anyone else* act in an event this
  individual was also present at (previously)?

Both are returned as instantaneous (most-recent-event) and cumulative
(running total) versions.

## A worked example

```{r toy-data}
dat <- data.frame(
  event = c(1, 1, 1,  2, 2, 2,  3, 3, 3, 3,  4, 4, 4, 4),
  id    = c(1, 2, 3,  1, 2, 4,  2, 3, 4, 5,  5, 1, 6, 2),
  fired = c(1, 1, 1,  0, 1, 1,  0, 1, 0, 1,  1, 0, 1, 1)
)
dat
```

Reading this as an event log: individuals `1, 2, 3` are all present (and
all "fire") at event 1; individuals `1, 2, 4` are present at event 2 (only
`2` and `4` fire); and so on. Rows do not need to arrive pre-sorted by
event -- `exposure_dyn()` sorts internally and returns results back in your
original row order.

```{r exposure}
counts <- with(dat, exposure_dyn(
  id_indiv  = id,
  id_events = event,
  actions   = fired,
  offset    = 0
))

cbind(dat, counts)
```

- `cumsum` is simply the individual's own running total of `actions`, in
  event order.
- `exposure_i` / `exposure_i_cum` are the *instantaneous* and *cumulative*
  **indirect** exposure -- whether/how many times someone *else* fired in
  an event this individual attended.
- `exposure_d` / `exposure_d_cum` are the *instantaneous* and *cumulative*
  **direct** exposure -- the individual's own prior firing history, as seen
  from a subsequent event.

For example, individual `2` fires at event 1, is present (but does not
fire) at event 3, and fires again at event 4. By event 3, their cumulative
direct exposure reflects the one time they'd already fired (at event 1);
by event 4 it's incremented again to reflect event 3's outcome.

## The `offset` argument

`offset` shifts *how far back* "previous" reaches, in units of events. The
default `offset = 1` looks at the single most recent event before the
current one; `offset = 0` (used above, to keep the toy example easy to read
by hand) instead folds the current event's own outcome in immediately
rather than lagging it by one step. In practice, use whichever offset
matches how quickly you believe exposure should "count": `offset = 1` is
the more common choice for a same-day/next-incident lag, while larger
offsets let you skip a fixed number of intervening incidents.

```{r offset-compare}
c0 <- with(dat, exposure_dyn(id, event, fired, offset = 0))[, "exposure_i"]
c1 <- with(dat, exposure_dyn(id, event, fired, offset = 1))[, "exposure_i"]
cbind(dat, exposure_i_offset0 = c0, exposure_i_offset1 = c1)
```

## Using exposure as a `clogit_perm()` covariate

The typical workflow is: compute exposure on your full event log, attach it
back to your officer-report data frame as a new column, then include it as
a covariate in `clogit_perm()` just like any other predictor (recall
`sim_events()`'s own `exposed` column, used throughout
`vignette("getting-started")`, is conceptually the same kind of variable).

```{r exposure-as-covariate}
x <- sim_events(nevents = 300, nofficers = 200, seed = 122)

exp_counts <- with(x, exposure_dyn(
  id_indiv  = officerid,
  id_events = incidentid,
  actions   = pointed000001,
  offset    = 1
))

x$exposure_i_cum <- exp_counts[, "exposure_i_cum"]

ans <- clogit_perm(
  nperm = 500,
  pointed000001 ~ female + years + exposure_i_cum + strata(incidentid),
  data  = x,
  ncpus = 2
)
print(ans)
```

Because `exposure_dyn()` is defined purely from the event log (`id_indiv`,
`id_events`, `actions`) it composes cleanly with any outcome column,
including any of the `nsims > 1` replicate outcomes discussed in
`vignette("simulating-data")`.
