---
title: "Computing Dynamic Exposure"
author: "Sima Najafzadeh"
date: today
format:
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Computing Dynamic Exposure}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r}
#| label: setup
library(mc3logit)
```

A recurring covariate in matched case-control studies of contagious
behavior (peer effects, "exposure to violence," etc.) is: *has this
individual been near this behavior before, and how much?* `exposure_dyn()`
computes that kind of running exposure count from a longitudinal event
log, correctly handling two things that are easy to get wrong by hand:

- **Temporal ordering** -- exposure must only look *backwards* in time.
- **No double counting** -- an individual's own action in the current event
  shouldn't count as exposure to themselves.

## The two kinds of exposure

Given a data frame with one row per (individual, event) pair and a binary
`action` indicator, `exposure_dyn()` distinguishes:

- **Direct exposure** -- did *this* individual act in a *previous* event?
- **Indirect exposure** -- did *anyone else* act in an event this
  individual was also present at (previously)?

Both are returned as instantaneous (most-recent-event) and cumulative
(running total) versions.

## A worked example

```{r}
#| label: toy-data
dat <- data.frame(
  event = c(1, 1, 1,  2, 2, 2,  3, 3, 3, 3,  4, 4, 4, 4),
  id    = c(1, 2, 3,  1, 2, 4,  2, 3, 4, 5,  5, 1, 6, 2),
  fired = c(1, 1, 1,  0, 1, 1,  0, 1, 0, 1,  1, 0, 1, 1)
)
dat
```

Reading this as an event log: individuals `1, 2, 3` are all present (and
all "fire") at event 1; individuals `1, 2, 4` are present at event 2 (only
`2` and `4` fire); and so on. Rows do not need to arrive pre-sorted by
event -- `exposure_dyn()` sorts internally and returns results back in your
original row order.

The whole log fits in one picture, which makes the exposure counts below
much easier to check by hand. Each column is an event; a filled dot means
that individual acted, a hollow one means they were present but did not:

```{r}
#| label: fig-timeline
#| fig-cap: "The toy event log: who was present at each event, and who acted."
#| fig-width: 7
#| fig-height: 4.2
blue <- "#2a78d6"; grid_col <- "#e1e0d9"
axis_col <- "#c3c2b7"; muted <- "#898781"

par(mar = c(4.2, 4.2, 3, 1))
plot(NA, xlim = c(.6, 4.4), ylim = c(.5, 6.5), axes = FALSE,
     xlab = "Event (time order)", ylab = "Individual",
     main = "Who was present, and who acted", col.lab = muted,
     col.main = "#0b0b0b", cex.main = 1.05)
abline(h = 1:6, col = grid_col, lwd = 1)
for (e in unique(dat$event))
  segments(e, min(dat$id[dat$event == e]), e, max(dat$id[dat$event == e]),
           col = grid_col, lwd = 1)

acted <- dat$fired == 1
points(dat$event[acted],  dat$id[acted],  pch = 19, col = blue, cex = 1.5)
points(dat$event[!acted], dat$id[!acted], pch = 21, col = muted,
       bg = "white", cex = 1.5, lwd = 1.5)
axis(1, at = 1:4, col = axis_col, col.axis = muted)
axis(2, at = 1:6, las = 1, tick = FALSE, col.axis = muted)
legend("topleft", legend = c("Acted", "Present, did not act"),
       col = c(blue, muted), pt.bg = c(blue, "white"), pch = c(19, 21),
       bty = "n", text.col = muted, cex = .85, horiz = TRUE)
```

The vertical connectors mark co-presence: individuals joined by a line were
at the same event, and so can expose one another.

```{r}
#| label: exposure
counts <- with(dat, exposure_dyn(
  id_indiv  = id,
  id_events = event,
  actions   = fired,
  offset    = 0
))

cbind(dat, counts)
```

- `cumsum` is simply the individual's own running total of `actions`, in
  event order.
- `exposure_i` / `exposure_i_cum` are the *instantaneous* and *cumulative*
  **indirect** exposure -- whether/how many times someone *else* fired in
  an event this individual attended.
- `exposure_d` / `exposure_d_cum` are the *instantaneous* and *cumulative*
  **direct** exposure -- the individual's own prior firing history, as seen
  from a subsequent event.

For example, individual `2` fires at event 1, is present (but does not
fire) at event 3, and fires again at event 4. By event 3, their cumulative
direct exposure reflects the one time they'd already fired (at event 1);
by event 4 it's incremented again to reflect event 3's outcome.

### Order invariance

Because the sort happens internally, shuffling the input rows cannot change
the answer -- a useful property when your event log arrives from a database
in arbitrary order:

```{r}
#| label: order-invariance
set.seed(42)
shuffled <- dat[sample(nrow(dat)), ]

counts_shuffled <- with(shuffled, exposure_dyn(
  id_indiv = id, id_events = event, actions = fired, offset = 0
))

# Put the shuffled result back in the original row order and compare
identical(
  unname(counts_shuffled[order(as.integer(rownames(shuffled))), ]),
  unname(counts)
)
```

## The `offset` argument

`offset` shifts *how far back* "previous" reaches, in units of events. The
default `offset = 1` looks at the single most recent event before the
current one; `offset = 0` (used above, to keep the toy example easy to read
by hand) instead folds the current event's own outcome in immediately
rather than lagging it by one step. In practice, use whichever offset
matches how quickly you believe exposure should "count": `offset = 1` is
the more common choice for a same-day/next-incident lag, while larger
offsets let you skip a fixed number of intervening incidents.

```{r}
#| label: offset-compare
c0 <- with(dat, exposure_dyn(id, event, fired, offset = 0))[, "exposure_i"]
c1 <- with(dat, exposure_dyn(id, event, fired, offset = 1))[, "exposure_i"]
cbind(dat, exposure_i_offset0 = c0, exposure_i_offset1 = c1)
```

## Using exposure as a `clogit_perm()` covariate

The typical workflow is: compute exposure on your full event log, attach it
back to your officer-report data frame as a new column, then include it as
a covariate in `clogit_perm()` just like any other predictor (recall
`sim_events()`'s own `exposed` column, used throughout the
[getting started](getting-started.html) vignette, is conceptually the same
kind of variable).

```{r}
#| label: exposure-as-covariate
x <- sim_events(nevents = 300, nofficers = 200, seed = 122)

exp_counts <- with(x, exposure_dyn(
  id_indiv  = officerid,
  id_events = incidentid,
  actions   = pointed000001,
  offset    = 1
))

x$exposure_i_cum <- exp_counts[, "exposure_i_cum"]
x$exposure_d_cum <- exp_counts[, "exposure_d_cum"]
```

Before using it as a covariate it is worth seeing how the measure actually
behaves over an individual's history. For the four busiest officers, both
kinds of exposure climb monotonically, and indirect exposure typically runs
ahead of direct -- you see peers act more often than you act yourself:

```{r}
#| label: fig-trajectories
#| fig-cap: "Cumulative direct and indirect exposure for the four busiest officers."
#| fig-width: 7
#| fig-height: 4.8
blue <- "#2a78d6"; orange <- "#eb6834"
grid_col <- "#e1e0d9"; axis_col <- "#c3c2b7"; muted <- "#898781"

busiest <- names(sort(table(x$officerid), decreasing = TRUE))[1:4]

par(mfrow = c(2, 2), mar = c(3.2, 3.2, 2.2, .8), oma = c(4.2, 2.6, 3, .4))
for (o in busiest) {
  d <- x[x$officerid == as.numeric(o), ]
  d <- d[order(d$incidentid), ]
  n <- seq_len(nrow(d))
  plot(NA, xlim = range(n),
       ylim = range(c(d$exposure_i_cum, d$exposure_d_cum), na.rm = TRUE),
       axes = FALSE, xlab = "", ylab = "",
       main = paste("Officer", o), col.main = muted, cex.main = .95)
  grid(nx = NA, ny = NULL, col = grid_col, lty = 1)
  lines(n, d$exposure_i_cum, col = blue,   lwd = 2)
  lines(n, d$exposure_d_cum, col = orange, lwd = 2)
  axis(1, col = axis_col, col.axis = muted, cex.axis = .8)
  axis(2, col = axis_col, col.axis = muted, cex.axis = .8, las = 1)
}

mtext("Cumulative exposure accumulates over an officer's incidents",
      outer = TRUE, cex = 1, col = "#0b0b0b", font = 2, line = .9)
mtext("Incident (in time order)", side = 1, outer = TRUE, col = muted,
      cex = .9, line = 1.5)
mtext("Cumulative exposure", side = 2, outer = TRUE, col = muted,
      cex = .9, line = 1)

par(fig = c(0, 1, 0, 1), oma = c(0, 0, 0, 0), mar = c(0, 0, 0, 0), new = TRUE)
plot(0, 0, type = "n", bty = "n", xaxt = "n", yaxt = "n", xlab = "", ylab = "")
legend("bottom", legend = c("Indirect (a peer had acted before)",
                            "Direct (this officer had acted before)"),
       col = c(blue, orange), lwd = 2, horiz = TRUE, bty = "n",
       text.col = muted, cex = .85, seg.len = 1.6, inset = c(0, .005))
```

With that in hand, the cumulative measure drops straight into the model:

```{r}
#| label: exposure-model
ans <- clogit_perm(
  nperm = 500,
  pointed000001 ~ female + years + exposure_i_cum + strata(incidentid),
  data  = x,
  ncpus = 2
)
print(ans)
```

Because `exposure_dyn()` is defined purely from the event log (`id_indiv`,
`id_events`, `actions`) it composes cleanly with any outcome column,
including any of the `nsims > 1` replicate outcomes discussed in the
[simulating data](simulating-data.html) vignette.

## A caution on interpretation

Exposure covariates built this way are *observational*: an officer whose
peers act often is also, typically, an officer working the kind of
assignment where acting is common. `exposure_dyn()` guarantees the
mechanics (no look-ahead, no self-counting), but it cannot separate peer
influence from shared context. Conditioning on the incident via `strata()`
absorbs some of that, which is one more reason the matched-set design is
worth the trouble here.
