---
title: "Getting Started with mc3logit"
author: "Sima Najafzadeh"
date: today
format:
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started with mc3logit}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
```

## The problem

`mc3logit` was built for **matched case-control data**, the kind you get when
several individuals experience the same event and only some of them "act."
The canonical example (and the one this package ships a simulator for) is a
police incident: several officers respond to the same call, and some subset
of them draw and point their firearm. Every incident is a matched set (a
*stratum*), and within a stratum we want to know which officer- and
event-level covariates predict the outcome.

`survival::clogit()` already fits this model via a conditional
(fixed-effects) logistic regression that conditions out the stratum. What
`mc3logit` adds is **permutation-based inference** on top of that fit:
instead of trusting the asymptotic standard errors from `clogit()` (which
can be unreliable with small strata or rare events), `clogit_perm()`
repeatedly shuffles the outcome *within* comparable strata and refits the
model, building an empirical null distribution for every coefficient. This
is the approach used by Ridgeway (2016) for exactly this kind of
matched-officer data.

## Installation

```r
# install.packages("devtools")
devtools::install_github("sima-njf/mc3logit_project", build_vignettes = TRUE)
```

```{r}
#| label: setup
library(mc3logit)
```

## Step 1: get some matched case-control data

If you don't have your own data yet, `sim_events()` will simulate a
plausible dataset: officers respond to incidents, and whether each officer
points their firearm is generated from a logistic model that depends on the
officer's sex, years of experience, whether they were exposed to a previous
incident, and the violence level of the event.

```{r}
#| label: sim-data
set.seed(1)
x <- sim_events(
  nevents   = 300,
  nofficers = 200,
  par_officer_female = -.5,
  par_officer_years  = -.5,
  par_exposure_event = 0,
  par_exposure_prev  = .5,
  par_event_violence = 1,
  seed = 122
)

head(x)
```

Every row is one officer's report for one incident. `incidentid` is the
matched-set identifier (the *stratum*), and `pointed000001` is the binary
outcome (did that officer point their gun during that incident).

A quick look at the matched-set structure, which is what makes this a
conditional-logit problem in the first place:

```{r}
#| label: strata-structure
#| collapse: true
per_incident <- table(x$incidentid)

data.frame(
  Statistic = c("Reports", "Incidents", "Officers",
                "Median officers per incident", "Share who pointed"),
  Value = c(
    nrow(x),
    length(unique(x$incidentid)),
    length(unique(x$officerid)),
    stats::median(as.numeric(per_incident)),
    round(mean(x$pointed000001), 3)
  )
)
```

Strata are small -- a handful of officers each -- which is precisely the
regime where asymptotic standard errors are least trustworthy.

## Step 2: fit the permutation-based conditional logit

`clogit_perm()` has the same model interface as `survival::clogit()` --
you write a formula with a `strata()` term for the matched-set id -- plus
one extra argument, `nperm`, for how many permutations to draw.

```{r}
#| label: fit
ans <- clogit_perm(
  nperm = 1000,
  pointed000001 ~ female + years + exposed + strata(incidentid),
  data  = x,
  ncpus = 2   # keep vignette builds fast/portable; raise this for real analyses
)
```

Internally this:

1. Fits the baseline model once with `survival::clogit()`.
2. Uses `find_candidates()` to figure out, for every officer/report, which
   other rows could plausibly be swapped with it (see the
   [permutation inference](permutation-inference.html) vignette for the
   matching rule).
3. Draws `nperm` permutations of the outcome with `permute()`, refits the
   model each time, and collects the coefficient estimates.
4. Turns the spread of those `nperm` coefficient estimates into two-sided
   permutation p-values.

## Step 3: read the results

```{r}
#| label: print
print(ans)
```

By default coefficients are reported as **odds ratios** (`odds = TRUE`);
pass `odds = FALSE` to `print()` to see the raw log-odds coefficients
instead. Stars follow the usual convention (`***` p <= .01, `**` p <= .05,
`*` p <= .1), where the p-value is the *permutation* p-value, not the
Wald p-value from `clogit()`.

```{r}
#| label: plot
plot(ans, odds = TRUE)
```

`plot.clogit_perm()` draws a forest plot of the coefficients (or odds
ratios) with their permutation-based confidence intervals, which is often
the fastest way to communicate results in a talk or paper.

## Step 4: extract what you need programmatically

`clogit_perm` objects support the usual accessor generics, so they play
nicely with other tools:

```{r}
#| label: accessors
#| collapse: true
coef(ans)        # point estimates, from the baseline clogit() fit
confint(ans)     # normal-approximation CI built from the permutation p-values
head(vcov(ans))  # covariance matrix estimated *from* the permutation draws
nobs(ans)        # number of observations used in the baseline fit
formula(ans)     # the model formula
```

Two flavors of confidence interval are available through the `which.`
argument of `confint()`:

- `which. = "coef"` (the default): a normal-approximation interval whose
  standard error is backed out from the permutation p-value (or, with
  `sigma_perm = TRUE`, from the empirical standard deviation of the
  permuted coefficients).
- `which. = "dist"` (anything other than `"coef"`): the raw empirical
  quantiles of the permutation distribution itself.

```{r}
#| label: confint-variants
#| collapse: true
confint(ans, which. = "coef")
confint(ans, which. = "coef", sigma_perm = TRUE)
```

## How to compare against the asymptotic answer

Because `clogit_perm()` keeps the baseline `survival::clogit()` fit in
`$fit`, you can put the permutation p-values next to the Wald p-values and
see directly where the two approaches disagree:

```{r}
#| label: wald-vs-perm
wald <- summary(ans$fit)$coefficients

data.frame(
  Term      = rownames(wald),
  Estimate  = round(wald[, "coef"], 3),
  Wald_p    = round(wald[, "Pr(>|z|)"], 4),
  Perm_p    = round(ans$pvals, 4),
  row.names = NULL
)
```

Where the two columns agree, the asymptotics were fine and the permutation
test simply confirms them. Where they diverge, it is usually a sign that
some stratum is small or nearly separable, and the permutation p-value is
the one to trust.

## Where to go next

- [How the permutation inference works](permutation-inference.html) --
  how `find_candidates()`/`permute()` build valid within-stratum
  permutations, and how the p-values and both interval types are computed.
- [Simulating matched case-control data](simulating-data.html) --
  the `sim_events()`/`sim_events2()` data-generating process.
- [Dynamic exposure covariates](dynamic-exposure.html) --
  building running exposure measures with `exposure_dyn()`.
- [Validating the method](validation.html) -- type-I error and power of
  the permutation test under a known truth.

## References

Knijnenburg, T. A., Wessels, L. F. A., Reinders, M. J. T., & Shmulevich, I.
(2009). Fewer permutations, more accurate P-values. *Bioinformatics*,
25(12), 161-168. <doi:10.1093/bioinformatics/btp211>

Ridgeway, G. (2016). Officer risk factors associated with police shootings:
a matched case-control study. *Statistics and Public Policy*, 3(1), 1-6.
<doi:10.1080/2330443X.2015.1129918>
