---
title: "Getting Started with mc3logit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mc3logit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## The problem

`mc3logit` was built for **matched case-control data**, the kind you get when
several individuals experience the same event and only some of them "act."
The canonical example (and the one this package ships a simulator for) is a
police incident: several officers respond to the same call, and some subset
of them draw and point their firearm. Every incident is a matched set (a
*stratum*), and within a stratum we want to know which officer- and
event-level covariates predict the outcome.

`survival::clogit()` already fits this model via a conditional
(fixed-effects) logistic regression that conditions out the stratum. What
`mc3logit` adds is **permutation-based inference** on top of that fit:
instead of trusting the asymptotic standard errors from `clogit()` (which
can be unreliable with small strata or rare events), `clogit_perm()`
repeatedly shuffles the outcome *within* comparable strata and refits the
model, building an empirical null distribution for every coefficient. This
is the approach used by Ridgeway (2016) for exactly this kind of
matched-officer data.

## Installation

```r
# install.packages("devtools")
devtools::install_github("sima-njf/mc3logit_project")
```

```{r setup}
library(mc3logit)
```

## Step 1: get some matched case-control data

If you don't have your own data yet, `sim_events()` will simulate a
plausible dataset: officers respond to incidents, and whether each officer
points their firearm is generated from a logistic model that depends on the
officer's sex, years of experience, whether they were exposed to a previous
incident, and the violence level of the event.

```{r sim-data}
set.seed(1)
x <- sim_events(
  nevents   = 300,
  nofficers = 200,
  par_officer_female = -.5,
  par_officer_years  = -.5,
  par_exposure_event = 0,
  par_exposure_prev  = .5,
  par_event_violence = 1,
  seed = 122
)

head(x)
```

Every row is one officer's report for one incident. `incidentid` is the
matched-set identifier (the *stratum*), and `pointed000001` is the binary
outcome (did that officer point their gun during that incident).

## Step 2: fit the permutation-based conditional logit

`clogit_perm()` has the same model interface as `survival::clogit()` --
you write a formula with a `strata()` term for the matched-set id -- plus
one extra argument, `nperm`, for how many permutations to draw.

```{r fit, cache=TRUE}
ans <- clogit_perm(
  nperm = 1000,
  pointed000001 ~ female + years + exposed + strata(incidentid),
  data  = x,
  ncpus = 2   # keep vignette builds fast/portable; raise this for real analyses
)
```

Internally this:

1. Fits the baseline model once with `survival::clogit()`.
2. Uses `find_candidates()` to figure out, for every officer/report, which
   other rows could plausibly be swapped with it (see the
   `vignette("permutation-inference")` vignette for the matching rule).
3. Draws `nperm` permutations of the outcome with `permute()`, refits the
   model each time, and collects the coefficient estimates.
4. Turns the spread of those `nperm` coefficient estimates into two-sided
   permutation p-values.

## Step 3: read the results

```{r print}
print(ans)
```

By default coefficients are reported as **odds ratios** (`odds = TRUE`);
pass `odds = FALSE` to `print()` to see the raw log-odds coefficients
instead. Stars follow the usual convention (`***` p <= .01, `**` p <= .05,
`*` p <= .1), where the p-value is the *permutation* p-value, not the
Wald p-value from `clogit()`.

```{r plot}
plot(ans, odds = TRUE)
```

`plot.clogit_perm()` draws a forest plot of the coefficients (or odds
ratios) with their permutation-based confidence intervals, which is often
the fastest way to communicate results in a talk or paper.

## Step 4: extract what you need programmatically

`clogit_perm` objects support the usual accessor generics, so they play
nicely with other tools:

```{r accessors}
coef(ans)           # point estimates, from the baseline clogit() fit
confint(ans)         # normal-approximation CI built from the permutation p-values
head(vcov(ans))       # covariance matrix estimated *from* the permutation draws
nobs(ans)             # number of observations used in the baseline fit
formula(ans)          # the model formula
```

Two flavors of confidence interval are available through the `which.`
argument of `confint()`:

- `which. = "coef"` (the default): a normal-approximation interval whose
  standard error is backed out from the permutation p-value (or, with
  `sigma_perm = TRUE`, from the empirical standard deviation of the
  permuted coefficients).
- `which. = "dist"` (anything other than `"coef"`): the raw empirical
  quantiles of the permutation distribution itself.

```{r confint-variants}
confint(ans, which. = "coef")
confint(ans, which. = "coef", sigma_perm = TRUE)
```

See `vignette("permutation-inference")` for the details of how the
p-values and both types of interval are constructed, and
`vignette("simulating-data")` / `vignette("dynamic-exposure")` for the two
data-generation helpers (`sim_events()`/`sim_events2()` and
`exposure_dyn()`) used to build the covariates above.

## References

Ridgeway, G. (2016). Officer risk factors associated with police shootings:
a matched case-control study. *Statistics and Public Policy*, 3(1), 1-6.
