---
title: "How the Permutation Inference Works"
author: "Sima Najafzadeh"
date: today
format:
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{How the Permutation Inference Works}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
```

```{r}
#| label: setup
library(mc3logit)
```

This vignette walks through the machinery behind `clogit_perm()`: how
candidate permutations are found, how the permutation itself is drawn, how
p-values are computed from the permutation distribution, and why that
matters for matched case-control data specifically.

## Why not just trust `clogit()`'s standard errors?

`survival::clogit()` estimates a conditional (fixed-effects) logistic
model and reports Wald-type standard errors based on the observed
information matrix. Those are asymptotically valid, but matched
case-control studies often have:

- **Many small strata** (e.g. incidents with 2-4 officers each), so the
  asymptotics kick in slowly.
- **Rare outcomes** within some strata (e.g. almost nobody points a
  firearm), which can make the likelihood surface poorly approximated by a
  quadratic near the MLE.

Permutation inference sidesteps both issues: instead of relying on a
large-sample approximation, it builds the null distribution of each
coefficient directly, by refitting the *same* model on data where the
outcome has been randomly reassigned in a way that respects the matching
structure. If the covariate truly has no effect, permuted refits should
produce coefficients "as extreme" as the observed one fairly often; if it
has a strong effect, the observed coefficient should sit in the tail of the
permutation distribution.

## Step 1: what's a valid permutation here?

You cannot just shuffle the outcome column globally -- that would break the
matched-set structure the conditional logit relies on (every stratum needs
the same *number* of "positive" cases it started with, or the comparison
stops being apples-to-apples). `find_candidates()` figures out, for every
row `i`, the set of other rows `j` it could trade outcomes with. For each
feature (column) `k` of the `features` matrix, define
`diff = features[i, k] - features[j, k]` (or `abs(diff)` when
`as_abs[k] = TRUE`); rows `i` and `j` are candidates for each other only if,
**for every feature**, `lower[k] <= diff <= upper[k]`. In other words,
`upper`/`lower` define an allowed *band* around zero for the (signed or
absolute) difference in each matching feature.

Inside `clogit_perm()` the matching feature is simply the (integer-coded)
stratum id, with `upper = lower = 0` and `as_abs = TRUE` -- i.e. "a row can
only trade outcomes with another row whose stratum id is identical to its
own." That's exactly what conditional logit needs: permutations that
shuffle *within* strata, never across them. Passing a matrix with more
columns (e.g. stratum id *and* a coarsened covariate) lets you additionally
restrict swaps to rows that also match on that covariate.

```{r}
#| label: find-candidates
strata <- c(1, 1, 1, 2, 2, 3, 3, 3, 3)
cand   <- find_candidates(
  features = cbind(as.integer(as.factor(strata))),
  upper    = 0,
  lower    = 0,
  as_abs   = TRUE
)
cand
```

`find_candidates()` returns, for each row, a zero-indexed vector of the
positions it could be swapped with (including itself). Rows `1:3` (stratum
1) only ever list each other; rows `6:9` (stratum 3) only ever list each
other, and so on.

## Step 2: drawing one permutation

Because `clogit_perm()` always matches on the stratum id exactly, its
candidate sets are *equivalence classes*: every row in a stratum is a
candidate for every other, and for nothing else. The correct draw is
therefore the simplest one -- a **uniform shuffle of the rows within each
stratum**, which is exactly what `clogit_perm()` does:

```{r}
#| label: permute-once
set.seed(1)
groups <- c(1, 1, 1, 2, 2, 3, 3, 3, 3)
idx    <- split(seq_along(groups), groups)

ord <- seq_along(groups)
for (g in idx) ord[g] <- sample(g)
ord
```

`clogit_perm()` draws one such shuffle per requested permutation
(`nperm`, default `1000`), reassigns the outcome column according to that
draw, and refits `survival::clogit()` on the permuted data. This is where
most of the runtime goes, which is why `clogit_perm()` exposes `ncpus` and
`parallel_args` to spread refits across a cluster.

::: {.callout-warning}
## `permute()` is not a uniform draw

The package also exports `permute()`, which builds a permutation from a
general `find_candidates()` structure by repeatedly pairing off mutual
candidates and swapping them. It is *not* uniform over the valid
permutations: `find_candidates()` never lists a row as its own candidate,
so a row gets swapped whenever any partner is still free, and the identity
is effectively unreachable.

For a stratum of exactly two rows this makes the draw **deterministic** --
the pair always trades places, where a uniform draw would leave it
untouched half the time. Since small strata dominate matched case-control
data, using `permute()` to build a reference distribution understates its
spread and yields anti-conservative p-values. `clogit_perm()` therefore
does not use it; see the [validation vignette](validation.html) for the
measured effect.
:::

### Sanity check: the permutation stays inside its stratum

The property the whole method rests on is that a permutation never moves an
outcome across matched sets, and never changes how many positives a stratum
has. Both are easy to verify directly:

```{r}
#| label: validity-check
#| collapse: true
set.seed(7)
grp <- rep(1:6, each = 4)
y   <- rep(c(1, 1, 0, 0), times = 6)
idx2 <- split(seq_along(grp), grp)

checks <- t(vapply(1:200, function(i) {
  p <- seq_along(grp)
  for (g in idx2) p[g] <- sample(g)
  c(
    stays_in_stratum   = all(grp[p] == grp),
    preserves_positives = isTRUE(all.equal(
      as.numeric(tapply(y[p], grp, sum)),
      as.numeric(tapply(y,    grp, sum))
    ))
  )
}, logical(2)))

colMeans(checks)   # both should be exactly 1
```

### Sanity check: the identity has to be reachable

A uniform shuffle of a stratum of size $m$ leaves any given row in place
with probability exactly $1/m$. That sounds like a technicality, but it is
the property whose absence broke the earlier implementation: a draw that
*always* moves every row is not a random permutation, it is a fixed
relabelling.

```{r}
#| label: fixed-point-check
#| collapse: true
set.seed(131)

fixed_point_rate <- function(m, draws = 2000) {
  g <- rep(1:60, each = m)
  ix <- split(seq_along(g), g)
  mean(replicate(draws, {
    p <- seq_along(g)
    for (s in ix) p[s] <- sample(s)
    mean(p == seq_along(g))
  }))
}

data.frame(
  stratum_size = c(2, 3, 4, 5),
  observed     = round(vapply(c(2, 3, 4, 5), fixed_point_rate, numeric(1)), 3),
  expected     = round(1 / c(2, 3, 4, 5), 3)
)
```

The observed and expected columns should agree. `permute()` returns `0.000`
for every even stratum size here, which is what the callout above is
warning about.

## Step 3: from permuted coefficients to p-values

For each permutation $b = 1, \ldots, B$, `clogit_perm()` records the
refit coefficient vector $\hat\beta^{(b)}$. The two-sided permutation
p-value for coefficient $j$ is then

$$
p_j = 2 \cdot \min\left(
  \frac{1}{B}\sum_{b=1}^B \mathbb{1}\{\hat\beta_j^{(b)} < \hat\beta_j\},\ \
  1 - \frac{1}{B}\sum_{b=1}^B \mathbb{1}\{\hat\beta_j^{(b)} < \hat\beta_j\}
\right)
$$

i.e., "how far into the tail of the permutation distribution does the
observed coefficient sit, doubled for a two-sided test." This is exactly
what `clogit_perm()` computes internally (see `pvals` in the returned
object).

### The zero p-value problem

A p-value of *exactly* zero would mean "not a single one of `nperm`
permutations was as extreme as the observed data" -- which is a statement
about your permutation *sample*, not a claim that the true p-value is
literally zero (the observed, unpermuted arrangement is itself one member
of the reference distribution, so the true p-value can never be exactly
zero). Following Knijnenburg et al. (2009), `mc3logit` replaces any
p-value that rounds to zero with the pseudo-count `1 / nperm`, which is the
smallest p-value that `nperm` permutations could actually resolve.

```{r}
#| label: pseudo-count
ans <- clogit_perm(
  nperm = 200,
  pointed000001 ~ female + years + exposed + strata(incidentid),
  data  = sim_events(300, 200, seed = 122),
  ncpus = 2
)
ans$pvals
```

If you need finer resolution near very small p-values, increase `nperm`
(the smallest resolvable p-value is always `1 / nperm`).

### Seeing the reference distribution

The permutation draws themselves are kept in `$coefs`, so the null
distribution the p-value is read off is directly inspectable:

```{r}
#| label: null-distribution
#| fig-height: 4
op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
for (v in c("female", "years")) {
  hist(
    ans$coefs[, v],
    breaks = 25,
    main   = paste("Permutation null:", v),
    xlab   = "coefficient",
    col    = "grey85",
    border = "white"
  )
  abline(v = coef(ans)[v], col = "firebrick", lwd = 2)
}
par(op)
```

The red line is the observed coefficient. The further into the tail it
falls, the smaller the permutation p-value -- and the shape of the
histogram tells you whether the normal approximation used by
`which. = "coef"` is reasonable in the first place.

## Step 4: two ways to build a confidence interval

`confint.clogit_perm()` offers two interval constructions, both built from
the same permutation draws but used differently:

1. **`which. = "coef"` (default)** -- treats the coefficient as
   approximately normal and *inverts* the permutation p-value to obtain an
   implied standard error, `sigma = |coef / qnorm(pval / 2)|`. This gives a
   symmetric interval centred on the observed coefficient, and is directly
   comparable to a Wald interval. Setting `sigma_perm = TRUE` instead uses
   `sqrt(diag(vcov(object)))` -- i.e., the actual empirical standard
   deviation of the `nperm` permuted coefficients -- as `sigma`.
2. **`which. != "coef"`** -- returns the raw empirical quantiles of the
   permutation distribution itself (e.g. the 2.5th/97.5th percentiles of
   $\hat\beta_j^{(1)}, \ldots, \hat\beta_j^{(B)}$), with no normality
   assumption at all.

```{r}
#| label: two-cis
ci_sd <- confint(ans, which. = "coef", sigma_perm = TRUE)
rbind(
  normal_from_pval   = confint(ans, which. = "coef")["female", ],
  normal_from_sd     = ci_sd["female", ],
  empirical_quantile = confint(ans, which. = "quantile")["female", ]
)
```

When the permutation distribution is close to normal, the three should
roughly agree; large discrepancies are themselves diagnostic of a skewed or
heavy-tailed reference distribution, in which case the empirical-quantile
interval is the more trustworthy of the two.

## Choosing `nperm`

`nperm` controls the resolution of the p-value, and nothing else: the
smallest value the test can report is `1 / nperm`. A rough guide:

| `nperm` | Smallest resolvable p | Reasonable use |
|--------:|----------------------:|:---------------|
| 200     | 0.005                 | Interactive exploration |
| 1,000   | 0.001                 | Default; fine for most reporting |
| 10,000  | 0.0001                | Publication, or p-values near a threshold |

Runtime is essentially linear in `nperm`, so raising it is cheap in wall
time if you also raise `ncpus`.

## References

Knijnenburg, T. A., Wessels, L. F. A., Reinders, M. J. T., & Shmulevich, I.
(2009). Fewer permutations, more accurate P-values. *Bioinformatics*,
25(12), 161-168. <doi:10.1093/bioinformatics/btp211>

Ridgeway, G. (2016). Officer risk factors associated with police shootings:
a matched case-control study. *Statistics and Public Policy*, 3(1), 1-6.
<doi:10.1080/2330443X.2015.1129918>
