---
title: "How the Permutation Inference Works"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How the Permutation Inference Works}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mc3logit)
```

This vignette walks through the machinery behind `clogit_perm()`: how
candidate permutations are found, how the permutation itself is drawn, how
p-values are computed from the permutation distribution, and why that
matters for matched case-control data specifically.

## Why not just trust `clogit()`'s standard errors?

`survival::clogit()` estimates a conditional (fixed-effects) logistic
model and reports Wald-type standard errors based on the observed
information matrix. Those are asymptotically valid, but matched
case-control studies often have:

- **Many small strata** (e.g. incidents with 2-4 officers each), so the
  asymptotics kick in slowly.
- **Rare outcomes** within some strata (e.g. almost nobody points a
  firearm), which can make the likelihood surface poorly approximated by a
  quadratic near the MLE.

Permutation inference sidesteps both issues: instead of relying on a
large-sample approximation, it builds the null distribution of each
coefficient directly, by refitting the *same* model on data where the
outcome has been randomly reassigned in a way that respects the matching
structure. If the covariate truly has no effect, permuted refits should
produce coefficients "as extreme" as the observed one fairly often; if it
has a strong effect, the observed coefficient should sit in the tail of the
permutation distribution.

## Step 1: what's a valid permutation here?

You cannot just shuffle the outcome column globally -- that would break the
matched-set structure the conditional logit relies on (every stratum needs
the same *number* of "positive" cases it started with, or the comparison
stops being apples-to-apples). `find_candidates()` figures out, for every
row `i`, the set of other rows `j` it could trade outcomes with. For each
feature (column) `k` of the `features` matrix, define
`diff = features[i, k] - features[j, k]` (or `abs(diff)` when
`as_abs[k] = TRUE`); rows `i` and `j` are candidates for each other only if,
**for every feature**, `lower[k] <= diff <= upper[k]`. In other words,
`upper`/`lower` define an allowed *band* around zero for the (signed or
absolute) difference in each matching feature.

Inside `clogit_perm()` the matching feature is simply the (integer-coded)
stratum id, with `upper = lower = 0` and `as_abs = TRUE` -- i.e. "a row can
only trade outcomes with another row whose stratum id is identical to its
own." That's exactly what conditional logit needs: permutations that
shuffle *within* strata, never across them. Passing a matrix with more
columns (e.g. stratum id *and* a coarsened covariate) lets you additionally
restrict swaps to rows that also match on that covariate.

```{r find-candidates}
strata <- c(1, 1, 1, 2, 2, 3, 3, 3, 3)
cand   <- find_candidates(
  features = cbind(as.integer(as.factor(strata))),
  upper    = 0,
  lower    = 0,
  as_abs   = TRUE
)
cand
```

`find_candidates()` returns, for each row, a zero-indexed vector of the
positions it could be swapped with (including itself). Rows `1:3` (stratum
1) only ever list each other; rows `6:9` (stratum 3) only ever list each
other, and so on.

## Step 2: drawing one permutation

`permute()` takes that candidate list and returns one permuted assignment
consistent with it. Under the hood it repeatedly (a) picks a not-yet-used
row, (b) picks one of its not-yet-used candidates at random, and (c) swaps
the two -- i.e. it builds a random **pairwise matching** among rows that
are mutual candidates, rather than one long cycle through the whole
stratum. A row that runs out of available candidates (for instance, the
odd one out in an odd-sized stratum, once everyone else has already been
paired off) is simply left unswapped for that draw.

```{r permute-once}
set.seed(1)
permute(cand) + 1L   # +1 to go back to R's 1-based indexing
```

`clogit_perm()` calls `permute()` once per requested permutation
(`nperm`, default `1000`), reassigns the outcome column according to that
draw, and refits `survival::clogit()` on the permuted data. This is where
most of the runtime goes, which is why `clogit_perm()` exposes `ncpus` and
`parallel_args` to spread refits across a cluster.

### Sanity check: is the matching reasonably well spread out?

A quick way to sanity-check `permute()` is to look at how often each row
ends up swapped with each of its candidates (versus staying put) over many
draws -- within a stratum, no single pairing should dominate:

```{r uniformity-check}
set.seed(131)
grp   <- sample(rep(1:20, each = 10))               # 20 strata of size 10
cand  <- find_candidates(cbind(as.integer(as.factor(grp))), 0, 0, TRUE)
perm  <- do.call(rbind, replicate(1000, permute(cand), simplify = FALSE))

# For row 1, what did it get paired with across 1000 draws?
# (its own index means "left unswapped that draw")
prop.table(table(perm[, 1] + 1L))
```

With 10 mutual candidates (including itself) in its stratum, no single
outcome should dominate the distribution above.

## Step 3: from permuted coefficients to p-values

For each permutation `b = 1, ..., nperm`, `clogit_perm()` records the
refit coefficient vector `\hat\beta^{(b)}`. The two-sided permutation
p-value for coefficient `j` is then

$$
p_j = 2 \cdot \min\left(
  \frac{1}{B}\sum_{b=1}^B \mathbb{1}\{\hat\beta_j^{(b)} < \hat\beta_j\},\ \
  1 - \frac{1}{B}\sum_{b=1}^B \mathbb{1}\{\hat\beta_j^{(b)} < \hat\beta_j\}
\right)
$$

i.e., "how far into the tail of the permutation distribution does the
observed coefficient sit, doubled for a two-sided test." This is exactly
what `clogit_perm()` computes internally (see `pvals` in the returned
object).

### The zero p-value problem

A p-value of *exactly* zero would mean "not a single one of `nperm`
permutations was as extreme as the observed data" -- which is a statement
about your permutation *sample*, not a claim that the true p-value is
literally zero (the observed, unpermuted arrangement is itself one member
of the reference distribution, so the true p-value can never be exactly
zero). Following Knijnenburg et al. (2009), `mc3logit` replaces any
p-value that rounds to zero with the pseudo-count `1 / nperm`, which is the
smallest p-value that `nperm` permutations could actually resolve.

```{r pseudo-count}
ans <- clogit_perm(
  nperm = 200,
  pointed000001 ~ female + years + exposed + strata(incidentid),
  data  = sim_events(300, 200, seed = 122),
  ncpus = 2
)
ans$pvals
```

If you need finer resolution near very small p-values, increase `nperm`
(the smallest resolvable p-value is always `1 / nperm`).

## Step 4: two ways to build a confidence interval

`confint.clogit_perm()` offers two interval constructions, both built from
the same permutation draws but used differently:

1. **`which. = "coef"` (default)** -- treats the coefficient as
   approximately normal and *inverts* the permutation p-value to obtain an
   implied standard error, `sigma = |coef / qnorm(pval / 2)|`. This gives a
   symmetric interval centred on the observed coefficient, and is directly
   comparable to a Wald interval. Setting `sigma_perm = TRUE` instead uses
   `sqrt(diag(vcov(object)))` -- i.e., the actual empirical standard
   deviation of the `nperm` permuted coefficients -- as `sigma`.
2. **`which. != "coef"`** -- returns the raw empirical quantiles of the
   permutation distribution itself (e.g. the 2.5th/97.5th percentiles of
   `\hat\beta_j^{(1)}, ..., \hat\beta_j^{(B)}`), with no normality
   assumption at all.

```{r two-cis}
ci_sd <- confint(ans, which. = "coef", sigma_perm = TRUE)
rbind(
  normal_from_pval   = confint(ans, which. = "coef")["female", ],
  normal_from_sd     = ci_sd["female", ],
  empirical_quantile = confint(ans, which. = "quantile")["female", ]
)
```

When the permutation distribution is close to normal, the three should
roughly agree; large discrepancies are themselves diagnostic of a skewed or
heavy-tailed reference distribution, in which case the empirical-quantile
interval is the more trustworthy of the two.

## References

Knijnenburg, T. A., Wessels, L. F. A., Reinders, M. J. T., & Shmulevich, I.
(2009). Fewer permutations, more accurate P-values. *Bioinformatics*,
25(12), 161-168. \doi{10.1093/bioinformatics/btp211}

Ridgeway, G. (2016). Officer risk factors associated with police shootings:
a matched case-control study. *Statistics and Public Policy*, 3(1), 1-6.
