---
title: "Simulating Matched Case-Control Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating Matched Case-Control Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mc3logit)
```

`mc3logit` ships two C++-backed simulators, `sim_events()` and
`sim_events2()`, that generate data with exactly the matched-set structure
`clogit_perm()` expects. They're useful for three things: learning the
package without needing real (often restricted) incident data, checking
that a model recovers known parameters, and running power / type-I-error
simulations.

## The data-generating process

`sim_events()` builds a synthetic police force and a sequence of incidents:

1. **Officers** are generated once: `Female ~ Bernoulli(0.5)`, years of
   experience `~ DiscreteUnif(min_year, max_year)`, and an "action rate"
   `~ Unif(min_rate, max_rate)` governing how quickly that officer reacts
   within an incident.
2. **Incidents** are generated one at a time. Each draws
   `nofficers ~ DiscreteUnif(min_per_event, max_per_event)` responding
   officers. Within the incident, officers react sequentially (a Poisson
   process driven by each officer's action rate), and whether an officer
   points their firearm is a logistic function of:
   - their sex (`par_officer_female`),
   - their years of experience (`par_officer_years`),
   - whether *anyone else* in the incident has already pointed
     (`par_exposure_event`),
   - whether the officer was exposed in a *previous* incident
     (`par_exposure_prev`), and
   - the incident's violence level (`par_event_violence`), plus an
     officer-level fixed effect (`par_officer_fe`).

Every row of the returned data frame is one officer's report for one
incident.

```{r sim-basic}
set.seed(1)
x <- sim_events(nevents = 1000, nofficers = 400)
str(x)
```

```{r sim-columns, echo = FALSE, results = "asis"}
knitr::kable(data.frame(
  column = c(
    "officerid", "female", "years", "fixed_effect", "incidentid",
    "violence_level", "response_time", "first", "exposed",
    "pointed000001, ..."
  ),
  meaning = c(
    "Officer id",
    "1 if the officer is female",
    "Years of experience",
    "Officer's latent propensity to point the gun",
    "Incident id (the *stratum* for clogit)",
    "Violence level of the incident",
    "Simulated response time of the officer",
    "1 if the officer was first to react",
    "1 if the officer was exposed in a previous incident",
    "Binary outcome(s): did the officer point the gun"
  )
))
```

## Tuning the effect sizes

Every `par_*` argument is a coefficient in the underlying logistic model,
so you can dial in exactly the effect sizes you want to test for:

```{r sim-tuned, cache=TRUE}
x <- sim_events(
  nevents             = 20000,
  nofficers           = 200,
  par_officer_female  = -.5,
  par_officer_years   = -.5,
  par_exposure_event  = -.5,
  par_event_violence  = 1,
  par_exposure_prev   = .25,
  par_officer_fe      = 1,
  seed                = 445
)
```

Because the data-generating parameters are known exactly, you can check
that a *full* model (one that also knows the latent `fixed_effect` and
`violence_level`, which a real analysis usually cannot observe) recovers
them:

```{r sim-recover}
ans <- glm(
  pointed000001 ~ -1 + female + years + exposed + I(-first) + fixed_effect +
    violence_level,
  data = x, family = binomial()
)
summary(ans)
```

## Multiple replicate outcomes: `nsims`

For power analyses or type-I-error checks you typically want *many*
independent draws of the outcome, holding the officers/incidents (and
hence covariates) fixed. Setting `nsims > 1` does exactly that: the officer
and incident structure is generated once, and `nsims` independent outcome
columns (`pointed000001`, `pointed000002`, ...) are appended.

```{r sim-nsims}
x <- sim_events(nevents = 200, nofficers = 150, nsims = 5, seed = 10)
grep("^pointed", colnames(x), value = TRUE)
```

This is the pattern you'd use to, for example, estimate the empirical
size of `clogit_perm()`'s permutation test under a true null: simulate many
replicate outcomes with `par_officer_female = 0` and check that the
fraction of replicates with `pvals["female"] <= .05` is close to `0.05`.

```{r type1-sketch, eval = FALSE}
reps <- sim_events(
  nevents = 300, nofficers = 200, nsims = 200,
  par_officer_female = 0,   # true null for "female"
  seed = 99
)

rejected <- sapply(grep("^pointed", colnames(reps), value = TRUE), function(outcome) {
  f <- as.formula(paste(outcome, "~ female + years + strata(incidentid)"))
  fit <- clogit_perm(nperm = 500, f, data = reps, ncpus = 1)
  fit$pvals["female"] <= .05
})

mean(rejected)  # should be close to 0.05
```

## Using your own officers and events: `sim_events2()`

`sim_events()` generates officer- and incident-level covariates for you.
If you already have a real (or externally simulated) roster of officers
and incidents and just want to simulate *reactions* on top of them, use
`sim_events2()` instead -- it takes the event/officer structure as input
rather than generating it:

```{r sim-events2}
set.seed(2)
n_reports <- 500

x2 <- sim_events2(
  event_id       = sample(1:100, n_reports, replace = TRUE),
  officer_id     = sample(1:60,  n_reports, replace = TRUE),
  officer_female = rbinom(n_reports, 1, .4),
  officer_years  = sample(0:15, n_reports, replace = TRUE),
  par_officer_female = -.3,
  par_officer_years  = -.2,
  seed = 2
)

head(x2)
```

This is particularly handy when you want to hold the *matching structure*
of a real dataset fixed (same officers responding to the same incidents in
the same group sizes) while simulating counterfactual outcome-generating
processes on top of it -- e.g. to build a study-specific power curve before
collecting or requesting real data.

## Next steps

See `vignette("getting-started")` for fitting `clogit_perm()` on simulated
data end to end, and `vignette("dynamic-exposure")` for computing exposure
covariates (like the `exposed` column above) from a raw event log.
