---
title: "Simulating Matched Case-Control Data"
author: "Sima Najafzadeh"
date: today
format:
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Simulating Matched Case-Control Data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
```

```{r}
#| label: setup
library(mc3logit)
```

`mc3logit` ships two C++-backed simulators, `sim_events()` and
`sim_events2()`, that generate data with exactly the matched-set structure
`clogit_perm()` expects. They're useful for three things: learning the
package without needing real (often restricted) incident data, checking
that a model recovers known parameters, and running power / type-I-error
simulations.

## The data-generating process

`sim_events()` builds a synthetic police force and a sequence of incidents:

1. **Officers** are generated once: `Female ~ Bernoulli(0.5)`, years of
   experience `~ DiscreteUnif(min_year, max_year)`, and an "action rate"
   `~ Unif(min_rate, max_rate)` governing how quickly that officer reacts
   within an incident.
2. **Incidents** are generated one at a time. Each draws
   `nofficers ~ DiscreteUnif(min_per_event, max_per_event)` responding
   officers. Within the incident, officers react sequentially (a Poisson
   process driven by each officer's action rate), and whether an officer
   points their firearm is a logistic function of:
   - their sex (`par_officer_female`),
   - their years of experience (`par_officer_years`),
   - whether *anyone else* in the incident has already pointed
     (`par_exposure_event`),
   - whether the officer was exposed in a *previous* incident
     (`par_exposure_prev`), and
   - the incident's violence level (`par_event_violence`), plus an
     officer-level fixed effect (`par_officer_fe`).

Every row of the returned data frame is one officer's report for one
incident.

```{mermaid}
%%| label: fig-dgp
%%| fig-cap: "How sim_events() builds a dataset."
flowchart TD
    A["<b>Officers</b><br/>sex, years of experience,<br/>action rate, fixed effect"]
    B["<b>Incidents</b><br/>violence level,<br/>1-5 responding officers"]
    A --> C["<b>Officer x incident</b><br/>one row per report"]
    B --> C
    C --> D{"Logistic model<br/>of the par_* coefficients"}
    D --> E["pointed = 1"]
    D --> F["pointed = 0"]
```

Officers are drawn once and reused across incidents, which is what creates
the repeated-measures structure `exposure_dyn()` later exploits.

```{r}
#| label: sim-basic
set.seed(1)
x <- sim_events(nevents = 1000, nofficers = 400)
str(x)
```

The matched sets are deliberately small -- that is the regime conditional
logit is designed for, and the one where asymptotic standard errors are
least dependable:

```{r}
#| label: fig-strata
#| fig-cap: "Distribution of matched-set sizes."
#| fig-width: 7
#| fig-height: 4
blue <- "#2a78d6"; grid_col <- "#e1e0d9"; muted <- "#898781"

sizes <- table(table(x$incidentid))

par(mar = c(4.2, 4.2, 3, 1))
bp <- barplot(
  as.numeric(sizes), names.arg = names(sizes), border = NA, col = blue,
  ylim = c(0, max(sizes) * 1.12),
  xlab = "Officers in the incident", ylab = "Number of incidents",
  main = "Matched-set size is uniform on 1-5 by default",
  col.axis = muted, col.lab = muted, col.main = "#0b0b0b", cex.main = 1.05
)
grid(nx = NA, ny = NULL, col = grid_col, lty = 1, lwd = 1)
barplot(as.numeric(sizes), border = NA, col = blue, add = TRUE, axes = FALSE)
text(bp, as.numeric(sizes), labels = as.numeric(sizes), pos = 3,
     col = muted, cex = .8, xpd = NA)
```

`min_per_event` and `max_per_event` control this range directly, so you can
generate anything from strictly 1:1 matched pairs to large sets.

```{r}
#| label: sim-columns
#| echo: false
knitr::kable(data.frame(
  column = c(
    "officerid", "female", "years", "fixed_effect", "incidentid",
    "violence_level", "response_time", "first", "exposed",
    "pointed000001, ..."
  ),
  meaning = c(
    "Officer id",
    "1 if the officer is female",
    "Years of experience",
    "Officer's latent propensity to point the gun",
    "Incident id (the *stratum* for clogit)",
    "Violence level of the incident",
    "Simulated response time of the officer",
    "1 if the officer was first to react",
    "1 if the officer was exposed in a previous incident",
    "Binary outcome(s): did the officer point the gun"
  )
))
```

## Tuning the effect sizes

Every `par_*` argument is a coefficient in the underlying logistic model,
so you can dial in exactly the effect sizes you want to test for:

```{r}
#| label: sim-tuned
x <- sim_events(
  nevents             = 10000,
  nofficers           = 200,
  par_officer_female  = -.5,
  par_officer_years   = -.5,
  par_exposure_event  = -.5,
  par_event_violence  = 1,
  par_exposure_prev   = .25,
  par_officer_fe      = 1,
  seed                = 445
)
```

Because the data-generating parameters are known exactly, you can check
that a *full* model (one that also knows the latent `fixed_effect` and
`violence_level`, which a real analysis usually cannot observe) recovers
them:

```{r}
#| label: sim-recover
ans <- glm(
  pointed000001 ~ -1 + female + years + exposed + I(-first) + fixed_effect +
    violence_level,
  data = x, family = binomial()
)
summary(ans)
```

Lining the estimates up against the values we asked for makes the recovery
check explicit:

```{r}
#| label: recovery-table
truth <- c(female = -.5, years = -.5, exposed = .25, violence_level = 1)
est   <- coef(ans)[names(truth)]

data.frame(
  Term     = names(truth),
  Truth    = as.numeric(truth),
  Estimate = round(as.numeric(est), 3),
  Error    = round(as.numeric(est) - as.numeric(truth), 3)
)
```

The estimates should land close to the values passed in; residual
discrepancy shrinks as `nevents` grows. Plotted with their confidence
intervals, the recovery is easier to judge at a glance:

```{r}
#| label: fig-recovery
#| fig-cap: "Estimated coefficients against the values used to generate the data."
#| fig-width: 7
#| fig-height: 4.5
orange <- "#eb6834"; axis_col <- "#c3c2b7"

se <- summary(ans)$coefficients[names(truth), "Std. Error"]
lo <- est - 1.96 * se
hi <- est + 1.96 * se
k    <- length(truth)
ypos <- rev(seq_len(k))

par(mar = c(4.2, 8.5, 3, 1))
plot(NA, xlim = range(c(lo, hi, truth)) + c(-.15, .15), ylim = c(.5, k + .8),
     axes = FALSE, xlab = "Coefficient (log-odds)", ylab = "",
     main = "Simulated effects are recovered", col.lab = muted,
     col.main = "#0b0b0b", cex.main = 1.05)
abline(v = pretty(range(c(lo, hi, truth))), col = grid_col, lwd = 1)
segments(lo, ypos, hi, ypos, col = blue, lwd = 2, lend = 1)
points(est,   ypos, pch = 19, col = blue,   cex = 1.1)
points(truth, ypos, pch = 4,  col = orange, cex = 1.2, lwd = 2)
axis(1, col = axis_col, col.axis = muted, lwd = 1)
axis(2, at = ypos, labels = names(truth), las = 1, tick = FALSE, col.axis = muted)
legend("topleft", legend = c("Estimate (95% CI)", "True value"),
       col = c(blue, orange), pch = c(19, 4), lwd = c(2, NA), lty = c(1, NA),
       bty = "n", text.col = muted, cex = .85)
```

Each interval should cover its orange cross. Where it does not, either the
sample is too small or the model is missing a term the simulator used.

## Multiple replicate outcomes: `nsims`

For power analyses or type-I-error checks you typically want *many*
independent draws of the outcome, holding the officers/incidents (and
hence covariates) fixed. Setting `nsims > 1` does exactly that: the officer
and incident structure is generated once, and `nsims` independent outcome
columns (`pointed000001`, `pointed000002`, ...) are appended.

```{r}
#| label: sim-nsims
x <- sim_events(nevents = 200, nofficers = 150, nsims = 5, seed = 10)
grep("^pointed", colnames(x), value = TRUE)
```

Because the covariates are shared across replicates, differences between
outcome columns are pure sampling variation in the outcome-generating
process -- which is exactly what you want when estimating the size or power
of a test. The [validation vignette](validation.html) uses this to measure
the empirical type-I error rate and power of `clogit_perm()` directly.

## Using your own officers and events: `sim_events2()`

`sim_events()` generates officer- and incident-level covariates for you.
If you already have a real (or externally simulated) roster of officers
and incidents and just want to simulate *reactions* on top of them, use
`sim_events2()` instead -- it takes the event/officer structure as input
rather than generating it:

```{r}
#| label: sim-events2
set.seed(2)
n_reports <- 500

x2 <- sim_events2(
  event_id       = sample(1:100, n_reports, replace = TRUE),
  officer_id     = sample(1:60,  n_reports, replace = TRUE),
  officer_female = rbinom(n_reports, 1, .4),
  officer_years  = sample(0:15, n_reports, replace = TRUE),
  par_officer_female = -.3,
  par_officer_years  = -.2,
  seed = 2
)

head(x2)
```

This is particularly handy when you want to hold the *matching structure*
of a real dataset fixed (same officers responding to the same incidents in
the same group sizes) while simulating counterfactual outcome-generating
processes on top of it -- e.g. to build a study-specific power curve before
collecting or requesting real data.

## Which simulator should I use?

| | `sim_events()` | `sim_events2()` |
|:--|:--|:--|
| Officer roster | generated for you | you supply it |
| Incident structure | generated for you | you supply it |
| Covariates | simulated | taken from your inputs |
| Typical use | learning, generic power studies | study-specific power, counterfactuals on a real design |

Both accept the same `par_*` coefficients and the same `nsims` and `seed`
arguments, so you can move between them without rewriting your analysis
code.

## Next steps

- [Getting started](getting-started.html) -- fitting `clogit_perm()` on
  simulated data end to end.
- [Dynamic exposure](dynamic-exposure.html) -- computing exposure
  covariates (like the `exposed` column above) from a raw event log.
- [Validating the method](validation.html) -- putting these simulators to
  work measuring type-I error and power.
