---
title: "Validating the Permutation Test"
author: "Sima Najafzadehkhoei"
date: today
format:
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Validating the Permutation Test}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
```

```{r}
#| label: setup
library(mc3logit)
```

A test is only worth using if it does what it claims: reject a true null
about as often as its nominal level says it should, and reject a false null
more often than that. Because `sim_events()` generates data from a *known*
model, we can measure both directly rather than taking them on faith.

This vignette does that, and reports one result that is good and one that
is a genuine limitation you should know about before using the package.

```{r}
#| label: study-size
n_rep  <- 40    # replicate outcome draws per scenario
n_perm <- 100   # permutations per fit
alpha  <- 0.05  # nominal level
```

The study is deliberately small so the vignette builds quickly; with
`r n_rep` replicates the Monte Carlo standard error on a rate near 0.05 is
about `r round(sqrt(0.05 * 0.95 / n_rep), 3)`, so read everything below as
"the right neighbourhood," not a precise estimate. Raising `n_rep` and
`n_perm` is what you would do for a real simulation study.

## The experiment

`sim_events(nsims = n_rep)` generates the officer/incident structure once
and then draws `n_rep` **independent outcome columns** on top of it. Every
replicate shares the same covariates and the same matched-set design and
differs only in the realized outcome -- exactly what is needed to estimate
the sampling distribution of a test statistic.

```{r}
#| label: helper
study <- function(par_female, par_years, seed) {

  dat <- sim_events(
    nevents            = 150,
    nofficers          = 100,
    nsims              = n_rep,
    par_officer_female = par_female,
    par_officer_years  = par_years,
    par_exposure_event = 0,
    par_exposure_prev  = 0,
    par_event_violence = 0,
    par_officer_fe     = 0,
    seed               = seed
  )

  outcomes <- grep("^pointed", colnames(dat), value = TRUE)

  res <- vapply(outcomes, function(outcome) {

    f <- stats::as.formula(
      paste(outcome, "~ female + years + strata(incidentid)")
    )

    fit <- try(
      suppressWarnings(
        clogit_perm(nperm = n_perm, f, data = dat, ncpus = 1)
      ),
      silent = TRUE
    )

    if (inherits(fit, "try-error"))
      return(c(perm = NA_real_, wald = NA_real_, beta = NA_real_, sd_perm = NA_real_))

    wald <- summary(fit$fit)$coefficients

    c(
      perm    = unname(fit$pvals["female"]),
      wald    = unname(wald["female", "Pr(>|z|)"]),
      beta    = unname(stats::coef(fit)["female"]),
      sd_perm = stats::sd(fit$coefs[, "female"])
    )

  }, numeric(4))

  as.data.frame(t(res))
}
```

## Part 1: type-I error under a complete null

Here nothing at all affects the outcome, so `female` is genuinely
irrelevant and the permutation is exact. This is the case the method
*must* get right.

```{r}
#| label: null-run
null_res <- study(par_female = 0, par_years = 0, seed = 4001)

data.frame(
  Test           = c("Permutation", "Wald"),
  Rejection_rate = round(c(mean(null_res$perm <= alpha, na.rm = TRUE),
                           mean(null_res$wald <= alpha, na.rm = TRUE)), 3),
  Nominal        = alpha
)
```

A sharper diagnostic than the rejection rate is whether the reference
distribution has the *right width*. Compare the average spread of the
permutation draws against the true sampling variability of the coefficient
across replicates -- the ratio should be close to 1:

```{r}
#| label: null-spread
data.frame(
  sd_true  = round(stats::sd(null_res$beta, na.rm = TRUE), 3),
  sd_perm  = round(mean(null_res$sd_perm, na.rm = TRUE), 3),
  ratio    = round(mean(null_res$sd_perm, na.rm = TRUE) /
                     stats::sd(null_res$beta, na.rm = TRUE), 3)
)
```

Under a true null, p-values should also be roughly **uniform** on
$[0, 1]$, not merely small 5% of the time:

```{r}
#| label: null-uniformity
#| fig-height: 4
op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

hist(
  null_res$perm,
  breaks = seq(0, 1, by = 0.1),
  main = "Permutation p-values\n(complete null)",
  xlab = "p-value", col = "grey85", border = "white"
)
abline(h = sum(!is.na(null_res$perm)) / 10, col = "firebrick", lwd = 2, lty = 2)

obs <- sort(null_res$perm[!is.na(null_res$perm)])
plot(
  stats::ppoints(length(obs)), obs,
  xlab = "Theoretical uniform quantile", ylab = "Observed p-value",
  main = "P-P plot (complete null)", pch = 19, col = "grey40"
)
abline(0, 1, col = "firebrick", lwd = 2)

par(op)
```

Permutation p-values are *discrete* -- with `nperm = `r n_perm`` they live
on a grid of width `1/n_perm`, which shows up as visible steps.

## Part 2: power under a real effect

```{r}
#| label: power-run
alt_res <- study(par_female = -1.2, par_years = 0, seed = 4002)

data.frame(
  Test           = c("Permutation", "Wald"),
  Rejection_rate = round(c(mean(alt_res$perm <= alpha, na.rm = TRUE),
                           mean(alt_res$wald <= alpha, na.rm = TRUE)), 3)
)
```

```{r}
#| label: power-hist
hist(
  alt_res$perm,
  breaks = seq(0, 1, by = 0.05),
  main = "Permutation p-values (true effect = -1.2)",
  xlab = "p-value", col = "grey85", border = "white"
)
abline(v = alpha, col = "firebrick", lwd = 2, lty = 2)
```

Under a real effect the p-values pile up near zero instead of spreading
out. The gap between this histogram and the flat one above is the power of
the test.

## Part 3: a limitation -- nuisance covariates with real effects {#sec-limitation}

Everything above held *all* other coefficients at zero. Now keep
`female` null but let `years` genuinely affect the outcome:

```{r}
#| label: nuisance-run
nuis_res <- study(par_female = 0, par_years = -0.5, seed = 4003)

data.frame(
  Scenario       = c("complete null", "years effect present"),
  Rejection_rate = round(c(mean(null_res$perm <= alpha, na.rm = TRUE),
                           mean(nuis_res$perm <= alpha, na.rm = TRUE)), 3),
  sd_ratio       = round(c(
    mean(null_res$sd_perm, na.rm = TRUE) / stats::sd(null_res$beta, na.rm = TRUE),
    mean(nuis_res$sd_perm, na.rm = TRUE) / stats::sd(nuis_res$beta, na.rm = TRUE)
  ), 3),
  Nominal        = alpha
)
```

The test **over-rejects** in the second row, and the `sd_ratio` column
shows why: the permutation reference distribution is too narrow relative to
the coefficient's real sampling variability.

This is not a bug in the shuffling -- it is intrinsic to permuting the raw
outcome. Shuffling `y` within strata destroys the association between the
outcome and *every* covariate, so the reference distribution is built under
the **complete** null ("nothing matters") rather than the **partial** null
actually being tested ("`female` doesn't matter, but `years` does"). When a
nuisance covariate carries real signal, the complete-null distribution is
too tight and p-values come out too small.

::: {.callout-important}
## Practical guidance

Treat `clogit_perm()` p-values as trustworthy when the other covariates in
the model are weak or the model is effectively univariate, and as
**anti-conservative** when a nuisance covariate carries substantial signal.
In the latter case, cross-check against the Wald p-value in `$fit` and
prefer the more conservative of the two.

The standard remedy is residual permutation -- Freedman & Lane (1983) or
ter Braak (1992) -- which permutes residuals from the model fitted under
the partial null instead of the raw outcome. That is not currently
implemented here.
:::

## Takeaways

- Under a **complete null** the test is well calibrated: p-values are
  roughly uniform and the rejection rate is near nominal.
- Under a **real effect** it has good power.
- With a **nuisance covariate carrying real signal** it is
  anti-conservative, for the structural reason described in
  [Part 3](#sec-limitation). This is a property of raw-outcome permutation,
  not of this implementation specifically.

::: {.callout-note}
## Historical note

Before version 0.1.0, `clogit_perm()` drew permutations with `permute()`,
which builds a pairwise matching and can never leave a row in place. For
the two-row strata that dominate matched case-control data that made the
draw deterministic, collapsing the reference distribution and pushing the
complete-null rejection rate to roughly 0.2 against a nominal 0.05.
`clogit_perm()` now uses a uniform within-stratum shuffle, which is what
Part 1 measures.
:::

## References

Freedman, D., & Lane, D. (1983). A nonstochastic interpretation of reported
significance levels. *Journal of Business & Economic Statistics*, 1(4),
292-298.

ter Braak, C. J. F. (1992). Permutation versus bootstrap significance tests
in multiple regression and ANOVA. In *Bootstrapping and Related
Techniques*, 79-85.

Knijnenburg, T. A., Wessels, L. F. A., Reinders, M. J. T., & Shmulevich, I.
(2009). Fewer permutations, more accurate P-values. *Bioinformatics*,
25(12), 161-168. <doi:10.1093/bioinformatics/btp211>

## Next steps

- [How the permutation inference works](permutation-inference.html) -- the
  construction being validated here.
- [Simulating matched case-control data](simulating-data.html) -- the
  `nsims` mechanism this vignette relies on.
