---
title: "Computing dynamic exposure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing dynamic exposure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r setup}
library(mc3logit)
```

## What is "exposure"?

In the peer-effects use-of-force study `mc3logit` was built for, an
officer's *exposure* captures whether they have recently witnessed a peer
take an action (e.g., pointing a weapon), either directly (co-present at
the same incident) or indirectly (through a peer of a peer). `exposure_dyn()`
computes this longitudinally, incident by incident, without double
counting individuals who appear at multiple incidents.

The function takes three parallel vectors -- an individual id, an event
(incident) id, and a binary action indicator -- and returns cumulative and
per-event exposure counts.

## A worked example

```{r data}
dat <- data.frame(
  event = c(1,1,1, 2,2,2, 3,3,3,3, 4,4,4,4),
  id    = c(1,2,3, 1,2,4, 2,3,4,5, 5,1,6,2),
  fired = c(1,1,1, 0,1,1, 0,1,0,1, 1,0,1,1)
)
dat
```

Here, four incidents (`event`) are attended by overlapping sets of
officers (`id`), each of whom either did (`fired = 1`) or did not
(`fired = 0`) take the action of interest.

```{r exposure}
counts <- with(dat, exposure_dyn(
  id_indiv  = id,
  id_events = event,
  actions   = fired,
  offset    = 0
))

cbind(dat, counts)
```

The result has five columns:

- `cumsum`: cumulative count of actions taken so far.
- `exposure_i` / `exposure_d`: indirect and direct exposure at that event.
- `exposure_i_cum` / `exposure_d_cum`: their cumulative counterparts.

## The `offset` argument

`offset` shifts the window used to compute exposure relative to the event
being scored, letting you look further back (or forward) in the sequence
of incidents. `id_events` need not already be sorted -- `exposure_dyn()`
orders internally by event id and returns results aligned to the original
row order.

```{r offset}
counts_lag1 <- with(dat, exposure_dyn(
  id_indiv  = id,
  id_events = event,
  actions   = fired,
  offset    = 1
))

cbind(dat, counts_lag1)
```

## Using exposure as a predictor

In practice, exposure measures like these are joined back onto the
simulated (or real) incident data and used as predictors in
`clogit_perm()`, capturing how a peer's prior behavior is associated with
an officer's own use-of-force decision -- see the
`vignette("mc3logit")` for a full modeling example.
