---
title: "Getting started with mc3logit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with mc3logit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
```

## Motivation

Matched case-control conditional logistic regression (`survival::clogit()`)
relies on asymptotic standard errors. When matched sets (strata) are small
-- as is common when strata are, e.g., officers co-present at the same
incident -- those asymptotic approximations can be unreliable. `mc3logit`
addresses this by permuting the outcome *within* strata to build an
empirical reference distribution for the model coefficients, following the
permutation approach used by Ridgeway (2016).

This vignette walks through the two core pieces of the package:

1. Simulating incident-level data with `sim_events()`.
2. Fitting a permutation-based conditional logit with `clogit_perm()`, and
   inspecting the result with `print()`, `plot()`, and `confint()`.

```{r setup}
library(mc3logit)
```

## Simulating data

`sim_events()` generates data similar to that used in the peer-effects
use-of-force study the package was built for. Officers are simulated with a
sex, years of experience, and a fixed effect capturing their individual
propensity to point their weapon. Incidents are simulated with a violence
level, and each officer present at an incident reacts (or not) following a
logistic model driven by their own features and by their exposure to peers'
prior behavior.

```{r simulate, cache=TRUE}
x <- sim_events(
  nevents   = 200,
  nofficers = 300,
  seed      = 122
)

head(x)
```

Each row is one officer's report for one incident (`incidentid`), with
`pointed000001` recording whether that officer pointed their weapon.

## Fitting the permutation-based conditional logit

`clogit_perm()` has the same formula interface as `survival::clogit()`: the
outcome, predictors, and a `strata()` term identifying the matched sets
(here, the incident). Internally it fits the baseline model with
`survival::clogit()`, then refits the model `nperm` times after permuting
the outcome within each stratum, building up an empirical distribution for
each coefficient.

```{r fit, cache=TRUE}
ans <- clogit_perm(
  nperm = 1000,
  pointed000001 ~ female + years + exposed + strata(incidentid),
  data  = x
)
```

### Inspecting the fit

The `print()` method reports odds ratios (or coefficients, with
`odds = FALSE`) together with permutation-based confidence intervals and
p-values:

```{r print}
print(ans)
```

`plot()` visualizes the same confidence intervals:

```{r plot}
plot(ans)
```

### Confidence intervals

`confint()` supports two flavors. `which. = "coef"` (the default) returns a
confidence interval for the point estimates, using the permutation p-value
to back out an implied standard error. Passing `which. = "perm"` instead
returns the empirical quantiles of the permutation distribution itself:

```{r confint}
confint(ans)
confint(ans, which. = "perm")
```

`coef()` and `vcov()` behave as usual, drawing on the baseline fit and the
permutation-based covariance of the coefficients, respectively:

```{r coef-vcov}
coef(ans)
vcov(ans)
```

## References

Ouellet, F., Hashimi, S., & Vega Yon, G. G. (2023). Peer effects in police
officers' use of force. *Journal of Quantitative Criminology*.
\doi{10.1007/s10940-022-09546-9}

Ridgeway, G. (2016). Officer risk factors associated with police shootings:
a matched case-control study. *Statistics and Public Policy*, 3(1), 1-6.
\doi{10.1080/2330443X.2015.1129918}

Knijnenburg, T. A., Wessels, L. F. A., Reinders, M. J. T., & Shmulevich, I.
(2009). Fewer permutations, more accurate p-values. *Bioinformatics*,
25(12), 161-168. \doi{10.1093/bioinformatics/btp211}
